---
title: "Sampling"
---

Follow along with the demo to learn how to sample data from a population using R and the Tidyverse. You should get started by loading the tidyverse in a new R block.

### Task 1: Load the tidyverse
```{r}
library(tidyverse)
```



### Task 2: Load 1,000 rows of data with the ages of people from 18-80

```{r}
set.seed(42)

population <- tibble(
  age=sample(18:80, 1000, replace=TRUE)
)

population
```


### Task 3: Sample 100 rows of data from the dataset
```{r}
sampled_data <- population |> slice_sample(n=20)

sampled_data
```


### Task 4: Calculate the mean age and standard deviation of the sample

```{r}
sample_stats <- sampled_data |> summarise(mean_age=mean(age), sd_age=sd(age))

sample_stats
```



### Task 5: Calculate the mean age and standard deviation of the population
```{r}
population_stats <- population |> summarise(mean_age=mean(age), sd_age=sd(age))

population_stats
```


### Task 6: Compare the sample and population statistics

Compare the means and standard deviations. Are they similar? Go back and re-run the sampling and statistics calculations with a smaller sample size. How do the results change?

They are very close, which means that our random sampling with 100 rows was good and actually produced accurate data. When I lower the sample size, the standard deviation gets larger, which is accurate because there are less data points so it's less like the population.  

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
