---
title: "JOUR405: Statistics for Journalists"
subtitle: "Midterm Exam"
name: "Elizabeth Faragi"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# JOUR405: Statistics for Journalists
## Midterm Exam

For this exam, you'll analyze several datasets using R and the statistical concepts we've covered in class. Load the tidyverse before beginning, then complete each task. Write your code in the provided blocks and answer the questions in complete sentences.

## Load the tidyverse

```{r}
library(tidyverse)

```

---

## Part 1: Restaurant Health Inspections (25 points)

The first dataset contains restaurant health inspection scores from Montgomery County. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv`

The dataset includes the name of the establishment, the number of points for critical and non-critical areas, the total points, maximum points possible, the compliance score, and the grade.

### Task 1: Read the data (2 points)

```{r}
 health_inspection <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv")
```

### Task 2: Calculate the mean, standard deviation, minimum, and maximum of compliance scores. Save the results in a new dataframe called `compliance_stats`. (2 points for each)

```{r}
compliance_stats <- health_inspection |> summarize(mean_compliance=mean(compliance_score), sd_compliance=sd(compliance_score),min_compliance=min(compliance_score), max_compliance=max(compliance_score))

compliance_stats
```

### Task 3: Create a histogram of the compliance scores with a vertical line showing the mean. Include a descriptive title and label the x and y axes using `labs()`. (10 points)

```{r}

health_inspection |> 
  ggplot() + 
  geom_histogram(aes(x = compliance_score), binwidth = 3) +
  geom_vline(aes(xintercept = mean(compliance_score)), color = "red", linetype = "dashed", size = 1) +labs(
    title = "Distribution of Compliance Scores",
    subtitle = "Red line = Mean Compliance Score",
    x = "Compliance Score",
    y = "Count",
    caption = "Health Inspection Scores from Montgomery County"
  )

```

### Task 4: Write 2–3 sentences interpreting what the standard deviation and histogram tell us about the distribution of compliance scores. What is newsworthy, and what would your next reporting step be? (5 points)

**The standard deviation for this dataset is high, at almost 6 data points, which leads me to believe that this data is pretty spread out around the mean. My histogram does in fact reflect this. As you can see, the bins continue to get closer and closer to lower numbers (the lowest being 42), even though the mean is higher, at 96. These low bins being so far from the mean reflect the large standard deviation. I'd like to know which stores have lower compliance scores (as well as what a compliance score even is & how it is graded) and why they were given low scores. I'd also look into the type of stores that got these lower scores and compare them to the stores that had higher compliance scores to try and figure out why they got so low.**

---

## Part 2: Daily Transit Ridership — Z-Scores and Sampling (30 points)

Your editor wants you to identify unusual days in WMATA's daily ridership data that might be worth investigating as story leads. The dataset contains daily bus and rail totals. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv`

```{r}
library(knitr)
library(lubridate)
```


### Task 5: Load the WMATA ridership data. (2 points)

```{r}
wmata_daily <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv")
```

### Task 6: Using `summarize()`, calculate the mean and standard deviation for both bus and rail ridership. Save the results in a dataframe called `wmata_stats`. (5 points)

```{r}
wmata_stats <- wmata_daily |> summarize(mean_bus=mean(bus), sd_bus=sd(bus),mean_rail=mean(rail),sd_rail=sd(rail))

wmata_stats

```

### Task 7: Using `mutate()`, add two new columns to `wmata_daily`: `rail_zscore` (the z-score for rail ridership) and `bus_zscore` (the z-score for bus ridership). Save the result as `wmata_zscores`. (8 points)

```{r}
wmata_zscores <- wmata_daily %>%
  mutate(
    mean_bus=mean(bus),
    sd_bus=sd(bus),
    mean_rail=mean(rail),
    sd_rail=sd(rail),
    rail_zscore = (total-mean_rail)/sd_rail,
    bus_zscore = (total-mean_bus)/sd_bus)
```

### Task 8: Filter `wmata_zscores` to show only days where the absolute value of `rail_zscore` is greater than 2 (unusually high or low rail ridership). Arrange by the absolute value of the z-score, with the most extreme first. How many unusual days are there? (8 points)

```{r}
unusual_wmata <- wmata_zscores %>%
  filter(abs(rail_zscore) > 2) %>%
  arrange(desc(abs(rail_zscore)))

unusual_wmata %>%
  select(date, rail, rail_zscore) %>%
  kable(digits = 2)

```

### Task 9: Using `set.seed()`, take a random sample of 91 days from `wmata_daily` (roughly 25% of the data). Calculate the mean and standard deviation for bus and rail ridership in the sample. How do the sample statistics compare to the population statistics you calculated in Task 6? Write 1–2 sentences comparing the values and explaining what this suggests about random sampling. (4 points)

```{r}
set.seed(19)

sample_91 <- wmata_daily |> sample_n(91)

sample_91_stats <- sample_91 |> summarize(mean_sample_bus=mean(bus), sd_sample_bus=sd(bus), mean_sample_rail=mean(rail),sd_sample_rail=sd(rail))

sample_91_stats
```

### Task 10: Write 2–3 sentences: What do the unusual days in your z-score analysis suggest as story leads? (3 points)

The unusual days in my z-score make sense, given that this sample has a standard deviation for bus and rail ridership is extrodinarily high. Initially, I was shocked seeing z-scores as high as 6, but with the huge standard deviation numbers, it makes sense. There is a lot of variability in this data, most likely being that some people take rails more days than others, some people take buses more often than rails, and some days people don't travel at all. 

---

## Part 3: Maryland Car Theft Rates (25 points)

Your editor has assigned you a story about car thefts in Maryland and wants you to find out which counties have the highest rates. The dataset contains car theft counts for Maryland counties in 2023 along with population figures. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv`

### Task 11: Load the car thefts data. (2 points)

```{r}
md_car_thefts <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv")
```

### Task 12: Using `mutate()`, add a column called `theft_rate` that calculates the car theft rate per 10,000 residents for each county. Save the result back to `md_car_thefts`. In 1–2 sentences, explain why a per-capita rate is more useful than raw counts for this comparison. (8 points)

```{r}
md_car_thefts <- md_car_thefts |> mutate(theft_rate=(thefts_2023 / population)*10000)
```

The rate per 10,000 residents is more helpful in this case because the population varies so widely from county to county. You can't compare car thefts from Baltimore County to Kent County because the population varies so much, so obviously the rate of car thefts is going to vary. You need to almost standardize the data.  


### Task 13: Calculate the median car theft rate and the total number of car thefts statewide. Show the counties that have rates above the median. (7 points)

```{r}
median_car_theft_rate = median(md_car_thefts$theft_rate)
total_car_theft = sum(md_car_thefts$thefts_2023)
```

### Task 14: Write 2–3 sentences describing what these calculations reveal about the distribution of car thefts in Maryland. What is the lede of a story based on your findings? (8 points)

The median car theft rate is about 9 car thefts per 10,000 people. The total is 23,871. What I'm seeing is Maryland has a lot of car thefts, regardless of what county you are in. It seems like that's a consistent crime. What I am fascinated about is certain places where I expected more car thefts (like Somerset County and Montgomery County, which have high populations) have lower rates. 

Lede: Baltimore City's car theft rate in 2023 is staggeringly different than every other county in Maryland, even counties with a larger population. It's rate is 123 per 10,000 residents, 34 more car thefts than Maryland's second highest crime rate county, at 89 per 10,000 people. 

## Part 4: Data Analysis Scenario (20 points)

A city parks department releases data showing that its 12 community swimming pools average 187 visitors per day during the summer season, with a standard deviation of 42 visitors. Your editor suspects that a handful of pools are dramatically underperforming and wants a story. You also learn that the department's figures are based on a random sample of 60 days per pool rather than a complete count.

Write 3–4 sentences for each of the following questions:

### Question A: Given that the mean is 187 and standard deviation is 42, which statistical measures would you prioritize using and why? (7 points)

I'd look into the standard deviation. I think that the standard deviation is really high in this case. I want to know which pools are furthest from the mean, so that would require some Z-score analysis. I'd also like to look into which days were randomly sampled (if they happened to be really warm days, sunny days, or rainy days)  


### Question B: How would you use z-scores to identify which pools are genuinely underperforming, and why is that more useful than simply ranking pools by raw attendance? (7 points)

Raw attendance is a lot of numbers and it's not going to standardize the data. You also need to figure out how many people go to certain pools (maybe based on proximity to their house, membership requirements, etc). Z-scores are going to identify which pools are "underperforming" and allows you to figure out which pools you're gonna need to look at. You're also gonna find out if there is even a story: if the pool with the lowest attendance has the lowest membership, then obviously they aren't underperforming because their population is smaller. 

### Question C: The department collected its attendance figures using a random sample of 60 days per pool rather than tracking every day. What questions would you ask about the sample before relying on these figures for a story? What concerns would you raise in your story? Use statistical terminology where appropriate. (6 points)

I want to know how they sampled: did they spin a wheel? Was it truly random? I also want to know which days were sampled so I can cross check with weather patterns on those days. If a pool was sampled on mostly rainy days by coincidence, then their random sample wasn't accurate to true attendance throughout the summer. I would also question why they didn't just track every day: this seems like a relatively easy measure to track throughout the summer and not as costly, so why not use the true population instead of a sample? 
---

### Submission Instructions
- Save your work frequently
- Make sure all code blocks run without errors
- Provide clear explanations for your choices

Good luck!
