---
title: "HW10: Significance Testing & Effect Size"
name: Elizabeth Faragi 
---

You previously calculated the mean, margin of error and confidence interval for monthly car thefts in Prince George's County. Your editor now wants to know: it _looks_ like car thefts are decreasing lately, but is that change real, and how big is it? In this assignment, you'll use significance testing and effect size to answer those questions. Load the tidyverse and the data to get started.

### Task 1: Load the tidyverse and the data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
# LOAD THE TIDYVERSE
```

```{r}

# Read the data
car_thefts <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/car_thefts_month.csv") |>
  arrange(month)  # Sort by date

```

### Task 2: Split the data and test for significance (4 points)

We'll split the monthly data into two halves, and then ask whether there's a statistically significant difference between these periods. To do that, we'll use a t-test with our alpha at 0.05. First, fill in the hypotheses, then replace the REPLACE_ME values in the code.

**Null Hypothesis:** Car thefts have not changed. 

**Alternative Hypothesis:** Car thefts have changed. 

```{r}

# Find the halfway point and add a column placing each month in the first or second halves

midpoint <- ceiling(nrow(car_thefts) / 2)
car_thefts <- car_thefts |>
  mutate(period = if_else(row_number() <= midpoint, "First Half", "Second Half"))

# Perform t-test comparing the two time periods
t_test_periods <- t.test(total ~ period, data = car_thefts)

# Display results
t_test_periods

```

**Answer this question:** What is the p-value? Is it less than 0.05? Can we reject the null hypothesis? What does this tell your editor?

The p-value is very very small. It is 2.14x10^-05. This is less than 0.05. Therefore, we can reject the null hypothesis. This tells our editor that the change is real and there has been an overall change in car thefts in the first half vs the second half. 

### Task 3: Visualize the comparison (4 points)

Let's see the two periods side by side with their confidence intervals. Replace the REPLACE_ME values.

```{r}
# Summarize by period
period_summary <- car_thefts |>
  group_by(period) |>
  summarize(
    mean_thefts = mean(total),
    sd_thefts = sd(total),
    months = n(),
    se_thefts = sd_thefts / sqrt(months),
    margin_error = qt(0.975, df = 32 - 1) * se_thefts,
    ci_lower = mean_thefts - margin_error,
    ci_upper = mean_thefts + margin_error
  )

period_summary |>
ggplot(aes(x = period, y = mean_thefts, fill = period)) +
  geom_col() +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +
  geom_text(aes(label = round(mean_thefts, 1)), vjust = -0.5, size = 4) +
  labs(
    title = "Comparing Car Thefts: First Half vs. Second Half",
    subtitle = paste0("p-value = ", round(t_test_periods$p.value, 5)),
    x = "Time Period",
    y = "Average Number of Car Thefts"
  ) +
  scale_fill_manual(values = c("First Half" = "#66c2a5", "Second Half" = "#fc8d62")) +
  theme_minimal() +
  theme(legend.position = "none")

```

**Answer this question:** Look at the error bars (confidence intervals) on the chart. Do they overlap? What does that tell you about whether the difference between the two periods is real?

They do not overlap. This tells me that the difference is real because the second half estimated average does not exist at all in the first half estimated average. Furthermore, the second half confidence interval is less than thr first half, which helps support the theory that car thefts are decreasing.  

### Task 4: Calculate the effect size (5 points)

We've established that the decline is statistically significant. But how large is it? Let's calculate Cohen's d to measure the effect size. For a two-sample comparison, we use the pooled standard deviation:

**Pooled SD** = sqrt(((n1 - 1) × sd1² + (n2 - 1) × sd2²) / (n1 + n2 - 2))

**Cohen's d** = (mean1 - mean2) / pooled SD

Replace the REPLACE_ME values. You'll need to pull the values from `period_summary`.

```{r}
# Extract the values from period_summary
first_half_stats <- period_summary |> filter(period == "First Half")
second_half_stats <- period_summary |> filter(period == "Second Half")

# Calculate pooled standard deviation
pooled_sd <- sqrt(
  ((first_half_stats$months - 1) * first_half_stats$sd_thefts^2 +
   (second_half_stats$months - 1) * second_half_stats$sd_thefts^2) /
  (first_half_stats$months + second_half_stats$months - 2)
)

# Calculate Cohen's d
cohens_d <- (first_half_stats$mean_thefts - second_half_stats$mean_thefts) / pooled_sd

# Interpret it
effect_label <- case_when(
  abs(cohens_d) < 0.2 ~ "Negligible",
  abs(cohens_d) < 0.5 ~ "Small",
  abs(cohens_d) < 0.8 ~ "Medium",
  TRUE ~ "Large"
)

tibble(
  `First half mean` = round(first_half_stats$mean_thefts, 1),
  `Second half mean` = round(second_half_stats$mean_thefts, 1),
  `Difference` = round(first_half_stats$mean_thefts - second_half_stats$mean_thefts, 1),
  `Cohen's d` = round(cohens_d, 3),
  `Interpretation` = effect_label
)
```

**Answer this question:** How would you describe this effect size? Compare it to the effect sizes we calculated in class (the response times activity had a small effect, the housing prices activity had a medium effect). What does a large effect size mean for the story you'd write about car thefts?

This effect size is large, which is different than what we've seen in class. Response time had a small effect, meaning that the new police system was doing something but not something big enough to continue funding. The medium effect of housing prices was good and helped show that maybe the new metro line had an effect, but that could also be from renovation and housing market going up. Having a large effect in this scenarios makes me more likely to write this article because we're seeing a big decrease of car thefts and that's like completely confirmed by data, which is really nice to see. 

### Task 5: Putting it all together (7 points)

You now have three pieces of evidence: the confidence intervals from last week, the significance test from this assignment, and the effect size.

1. In Part 1, your editor asked "just tell me if car thefts are going up or down". Now you have significance testing. Write 2-3 sentences telling your editor the answer, using all three pieces of evidence (CI, p-value, and effect size).

The confidence interval (which predicts the average # of car thefts for a specific population) from the second half of the months does not overlap with the confidence interval of the first half, meaning that car thefts are changing and are not the same as they were from the first half of the months. The second half confidence interval is also less than the first half, so we have evidence that car thefts are, on average, decreasing. The p-value we found is very small, meaning that these results are unlikely to happen by chance. We also found that the effect size was large, so we can actually see that car thefts are decreasing by a lot and there is a big change lately in car thefts.  

2. Write a headline and a 2-3 sentence lede for a story about car thefts in Prince George's County. Unlike the response times story from class (where the effect was small and the headline needed to be cautious), this is a case where the data strongly supports a clear conclusion. Your writing should reflect that — but still note the limitations (correlation vs. causation, what else might explain the decline).

HEADLINE: PG COUNTY CAR THEFTS DECREASING 

Lede: Over the past few months, car thefts have been, on average, decreasing significanly. according to new data. 
Data analysis finds that the average # of car thefts per month since November 2024 falls between 279 and 345, a significant change. The average from July 2023 to October 2024 falls between 399 and 479, according to a t-test. 

When you are finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
