---
title: "HW3 WH Salaries"
---

In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2025.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2025.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
wh_salaries <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/white_house_salaries_2025.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

The first thing I noticed was that most of the staff members have pretty high salaries. Depening on the job, they range from 60,000 - almost 200,000 per year. However, there are a few staff members who have a salary of 0, which will really mess up the data when calculating averages. 

### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
wh_salaries |> summarize(mean_salary=mean(SALARY))
wh_salaries |> summarize(median_saLary=median(SALARY))
```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
wh_salaries |> summarize(mean_salary=mean(SALARY))
wh_salaries |> summarize(median_saLary=median(SALARY))
wh_salaries |> summarize(min_salary=min(SALARY))
wh_salaries |> summarize(max_salary=max(SALARY))
```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

The mean and median salaries are similar, but because the median is larger than the mean I know that the data is skewed left. In addition, the minimum salary is 0 and the highest is 225,700, both of which are extremely different from the mean and median. This leads me to believe that both these values are outliers of some sort, which the minimum salary being more of an outlier and pulling the mean down to an unrealistic average. 

### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

I'd like to look into the jobs that have a $0 salary. What do they do and why are they unpaid? Should they be paid? I feel like its odd to me that these people are working for free, but maybe I'm just unaware of how white house jobs work. I'm also curious of what the maximum salary does: it just says advisor, so I'm curious exactly what she does to make so much more than everyone else. 
