---
title: "HW8b: Polling Simulation"
---

In class, you simulated polling estimates for the share of Americans who are religiously unaffiliated — a group that makes up about 36% of the population. In this assignment, you'll run the same simulation for a different group: White Evangelicals, who make up roughly 10% of the population.

That difference matters more than it might seem. When a proportion is close to 50%, samples tend to be pretty well-behaved. When it's much smaller — or much larger — samples can behave differently, especially at small sizes. Your job is to run the simulation, compare what you find to the in-class results, and think carefully about what it means for reporting on religion and politics.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```

### Task 2: Load the CES data

Load the CES 2024 religion dataset from the same URL as the in-class activity and save it to a dataframe called `ces`:

`https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/ces_2024_religion.csv`

```{r}
ces <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/ces_2024_religion.csv")

```

### Task 3: Calculate the population proportion

Create a new column called `evangelical` that is `TRUE` if a respondent's `trad` value is `"White Evangelical"` and `FALSE` otherwise. Then calculate the proportion of the full dataset that is White Evangelical and save it as `population_proportion`.

```{r}

ces <- ces |>
  mutate(evangelical = trad %in% c("White Evangelical"))

population_proportion <- mean(ces$evangelical)

```

Write down this value. How does it compare to the unaffiliated proportion (about 0.36) from the in-class activity?

This value is about 0.105. This is much smaller than the unaffiliated proportion, but it's still a good chunk of people. 

### Task 4: Take a single small sample

Use `set.seed(42)` and sample 50 rows from `ces`. Save it as `sample_50`. Calculate the proportion of White Evangelical respondents in your sample.

```{r}
set.seed(42)

sample_50 <- ces |> sample_n(50)

proportion_50 <- mean(sample_50$evangelical)


```

How close is your sample estimate to the population proportion? How does the accuracy compare to what you saw with the unaffiliated group in class?

This sample has a proportion of 0.16. That's pretty close to the population proportion of 0.105, but it's larger. During class, my sample was pretty close to the unaffiliated population proportion, but it was still off, similar to how this sample compares to the true population proportion of evangelicals. 

### Task 5: Take a single larger sample

Use `set.seed(1966)` and sample 500 rows. Save it as `sample_500`. Calculate the proportion of White Evangelical respondents in this sample.

```{r}

set.seed(1966)

sample_500 <- ces |> sample_n(500)

proportion_500 <- mean(sample_500$evangelical)


```

Did the larger sample improve the estimate? How does the improvement compare to what you saw in class?

It definitely improved the estimate. A sample of 50 was off by about 0.55 points, while a sample of 500 is off by only 0.015. This tracks, as I saw in class that using a larger sample leads to estimates closer to the actual true population proportion. 

### Task 6: Simulate many polls

The code below runs the same simulation as the in-class activity, but using your `evangelical` column. Run it as-is.

```{r}
set.seed(42)

simulate_polls <- function(data, sample_size, n_simulations = 1000) {
  proportions <- replicate(n_simulations, {
    s <- data |> sample_n(sample_size)
    mean(s$evangelical)
  })
  tibble(proportion = proportions, sample_size = sample_size)
}

simulations_50 <- simulate_polls(ces, 50)
simulations_500 <- simulate_polls(ces, 500)
```

### Task 7: Visualize the simulation results

Create two histograms side by side using the code structure from class. Add a vertical line at `population_proportion`. Make sure to update the title and subtitle to reflect that this simulation is about White Evangelicals.

```{r}
simulations_all <- bind_rows(simulations_50, simulations_500) |> 
  mutate(sample_size = paste("n =", sample_size))  

ggplot(simulations_all, aes(x = proportion)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  geom_vline(xintercept = population_proportion, color = "red", linetype = "dashed", linewidth = 1) +
  facet_wrap(~sample_size) +
  labs(
    title = "Distribution of Sample Estimates: Share Religiously White Evangelical",
    subtitle = "Red line = true population proportion. Each histogram shows 1,000 simulated polls.",
    x = "Sample Proportion",
    y = "Count",
    caption = "Source: 2024 Cooperative Election Study"
  ) +
  theme_minimal()

```

### Task 8: Reflect on what you see

**Describe the shape of the n=50 histogram.** Is it as symmetric as the unaffiliated histogram from class? Why do you think that is, given that White Evangelicals make up only about 10% of the population?

It is symmetric, but not as symmetic as the unaffiliated histogram. This makes sense, as there is just less White Evangelicals, so their data is not going to be as cohesive since there is fewer of them. I will say that it is still relatively normal shaped. 

**Compare the n=50 and n=500 histograms for this group.** How much does the distribution change with a larger sample? Is the improvement bigger or smaller than what you saw in class with the unaffiliated group?

Again, we see with the larger sample size that the distribution gets tighter and more clustered around the true population proprtion. I think that this improvement is bigger compared to the unaffiliated group, most likley because the population of White Evangelicals is smaller. The bigger a sample size, the closer we're getting to the true population. Relatively speaking, if there are 1000 White Evangelicals and 5000 unaffiliated, a sample size of 500 is closer to the true population of Evangelicals, so that sample size is going to get much closer to the true propotion compared to the unaffilited one because 500 is just closer to 1000 than it is to 5000. 

**A news organization wants to report on White Evangelical voters using a poll of 200 people.** Based on your simulations, roughly how many White Evangelical respondents would that poll be likely to include? What does that suggest about the reliability of estimates for smaller subgroups?

If only about 10% of the population is White Evangelicals, only 20 of those people would be white evangelicals. This leads me to believe that this poll wouldn't be accurate for white evangelicals because this is such a small sample of their people, therefore producing less accurate results. Estimates for smaller subgroups are probably not that accurate when it comes to polling small samples. If the subgroup is already small and you poll a small sample, it's not going to get any bigger. 

**Imagine you are covering the 2028 presidential election and a campaign shares a poll showing White Evangelical support at 78%.** The poll sampled 400 people. Based on everything you've seen in this assignment and in class, what would you want to know before reporting that number?

How many of those 400 people were White Evangelicals, what questions were asked, how many other samples were taken (were there other samples taken), how did we calculte 78%. 

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
